Errorless irrationality: removing error-driven components from the inverse base-rate effect paradigm

### 2023-01-10 docs/manuscript.tex
broaden discussion

In this study, we tested a central assumption of the most prominent theories of the IBRE.
This central assumption was prediction error.
In our first attempt, we implemented an observational learning version of the canonical IBRE procedure.
This meant that features and category labels appeared on the screen at the same time.
Participants learned about categories by reading complete sentences that describe what symptoms people exhibit and what diseases they have.
The experiment included no feedback and required no responses from participants during training.
From a theoretical perspective, there was no opportunity for making an explicit error, therefore the condition for prediction error.
Regardless, we observed the inverse base-rate effect.
One limitation of this approach was that there are assumed causal relationships between features (symptoms) and labels (diseases).
These relationships can predispose participants to make feature-to-label predictions, which will result in prediction error and attentional relocation.
In our second attempt, we further removed the causal relationship between features and labels by changing the stimuli and their presentation.
Here, participants saw nothing but an arrangement of geometric shapes, where previous category labels were treated as features.
There were no causal links between features and labels.
When participants needed to complete incomplete arrangements of these shapes, they still exhibited a rare bias on $BC$ trials.
We still observed the IBRE.

@@

% LD: so it seems that Johansen et al treated asymmetric representation as an experimental condition and not as a theoretical construct
% LD: I think this is a problem because asymmetric representation is hypothesized to be an asymmetry in the way the two categories represented
Compared to \citeA{johansen2007paradoxical}, we do not consider the asymmetric representation and the base-rate neglect as necessary conditions.
From our point of view, they are theoretical constructs.
They are not part of the experimental design.
We consider them to result from the precise environmental conditions required for the IBRE as opposed to being environmental conditions necessary/sufficient for the IBRE.

@@

In addition, Experiment 2 suggests that the IBRE generalises beyond simple predictive-learning \cite{kruschke1996base,don2021hearing} and decision-making \cite{johansen2007paradoxical} paradigms.

>>>>

In this study, we tested a central assumption of the most prominent theories of the IBRE.
This central assumption was prediction error.

In our first attempt, we implemented an observational learning version of the canonical IBRE procedure.
This meant that features and category labels appeared on the screen at the same time.
Participants learned about categories by reading complete sentences that describe what symptoms people exhibit and what diseases they have.
The experiment included no feedback and required no responses from participants during training.
From a theoretical perspective, there was no opportunity for making an explicit error, therefore the condition for prediction error.
Regardless, we observed the inverse base-rate effect.
One limitation of this approach was that there are assumed causal relationships between features (symptoms) and labels (diseases).
These relationships can predispose participants to make feature-to-label predictions, which will result in prediction error and attentional relocation.

In our second attempt, we further removed the causal relationship between features and labels by changing the stimuli and their presentation.
Here, participants saw nothing but an arrangement of geometric shapes, where previous category labels were treated as features.
There were no causal links between features and labels.
When participants needed to complete incomplete arrangements of these shapes, they still exhibited a rare bias on $BC$ trials.
We still observed the IBRE.

In both experiments, the IBRE occured without any explicit detail in the experimental procedure that would result in prediction error.
Therefore, any theorised error-driven process must be able to operate without explicit feedback.
Most prominent theories and their corresponding formal specification rely on relocating attention in response to prediction error.
They are unable to process the current experiment, because they are not designed to encode information presented without feedback or fit cued recall tasks.

@@

% LD: so it seems that Johansen et al treated asymmetric representation as an experimental condition and not as a theoretical construct
% LD: I think this is a problem because asymmetric representation is hypothesized to be an asymmetry in the way the two categories represented
Compared to \citeA{johansen2007paradoxical}, we do not consider the asymmetric representation and the base-rate neglect as necessary conditions.
From our point of view, they are theoretical constructs.
They are not part of the experimental design.
We consider them to result from the environmental conditions specified in the experiment required for the IBRE as opposed to being necessary/sufficient conditions for the IBRE.

@@

In Experiment 1, we conducted a successful conceputal replication of \citeA{johansen2007paradoxical}, which gave evidence for the IBRE being independent of supervised learning procedures.
In addition, Experiment 2 further suggests that the IBRE generalises beyond simple predictive-learning \cite{kruschke1996base,don2021hearing} and decision-making \cite{johansen2007paradoxical} paradigms.
Theories of IBRE are inadequate to account for these findings, because of their inability to extend beyond supervised learning.
